|

### 2023-05-15 paper/01-introduction.Rmd
more results: players on the same team
more minor edits

In this manuscript, we specifically focus on measuring the performance of defensive linemen in the NFL. There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure \ref{fig:formation} (top) for example formation with DTs and DEs). The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl]. Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage (see Figure \ref{fig:formation} (bottom) for example defensive formation with a NT). Typically, teams will employ either one NT or two DTs on the interior with DEs along either side of the defensive line. In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) who typically play pass coverage.

@@

Perhaps the most commonly-known statistics for evaluating defensive linemen on pass rush plays are sacks, hits, and hurries, which are discretely observed at the play-level.
Officially tracked by the NFL since 1982, a sack is recorded when a defender tackles the QB behind the line of scrimmage before the QB releases a pass. 
Other traditional box score statistics such as hits and hurries are collected by various outlets. 
A hit is a collision between a defender and the opposing team's quarterback while the QB is in possession of the ball.
A hurry represents an instance when a defender successfully disrupts without necessarily making direct contact with the QB and forces the QB to make a throw earlier than expected.
These are all simple binary measures of pass rush outcome for any given play.
However, for plays that do not result in the aforementioned outcomes (in particular, sack), there are still many intermediate defensive actions on the field within the play that are valuable and can be considered successes.

Moreover, the sum of sacks, hits, and hurries is often defined as pressures.
This is better than sacks to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
Pro football focus (PFF) comes up with a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
However, the choice of weights is ad-hoc and still only indicates binary outcomes, similar to the shortcomings of previous metrics.

@@

@Wolfson2017Forecasting comment on the two aforementioned articles that "[a]lthough these are exciting preliminary steps, there is still a long way to go before we can provide a comprehensive appraisal of the achievements of an individual lineman."
The challenge here is fundamental, since there were not enough public data at the time to develop any meaningful metric for linemen in football, as also noted by @AlamarGould2008.
However, with the granularity of player tracking data, we have access to data not only for the linemen but also for every player on the field.
This provides us with a great opportunity to assess the performance of linemen in football.

>>>>

Our focus of this manuscript is specifically on measuring the performance of defensive linemen in the NFL. There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure \ref{fig:formation} (top) for example formation with DTs and DEs). The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl]. Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage (see Figure \ref{fig:formation} (bottom) for example defensive formation with a NT). Typically, teams will employ either one NT or two DTs on the interior with DEs along either side of the defensive line. In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) who typically play pass coverage.

@@

Perhaps the most commonly-known statistics for evaluating defensive linemen on pass rush plays are sacks, hits, and hurries, which are discretely observed at the play-level.
Officially tracked by the NFL since 1982, a sack is recorded when a defender tackles the QB behind the line of scrimmage before the QB releases a pass. 
Other traditional box score statistics such as hits and hurries are collected by various outlets. 
A hit is a collision between a defender and the opposing team's quarterback while the QB is in possession of the ball.
A hurry represents an instance when a defender successfully disrupts without necessarily making direct contact with the QB and forces the QB to make a throw earlier than expected.
These are all simple binary measures of pass rush outcome for any given play.
However, for plays that do not result in the aforementioned outcomes (in particular, sack), there are still many intermediate defensive actions on the field within the play that are valuable and can be considered positive achievements.

Moreover, the sum of sacks, hits, and hurries is often defined as pressures.
This is better than the individual counts to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
Pro football focus (PFF) comes up with a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
However, the choice of weights is ad-hoc and still only indicates binary outcomes, similar to the shortcomings of previous metrics.

@@

@Wolfson2017Forecasting comment on the two aforementioned articles that "[a]lthough these are exciting preliminary steps, there is still a long way to go before we can provide a comprehensive appraisal of the achievements of an individual lineman."
The challenge here is fundamental, since there were not enough public data at the time to develop any meaningful metric for linemen in football, as also noted by @AlamarGould2008.
However, with the granularity of player tracking data, we have access to data not only for the linemen but also for every player on the field.
This provides us with a great opportunity to study and gain better insights into linemen performance in football.

### 2023-05-15 paper/02-data.Rmd
more minor edits

Along with the tracking information, the 2023 BDB includes scouting data provided by Pro Football Focus (PFF). 
This contains manually collected player-level information at the play level, such as the player's role (e.g., whether they are a pass rusher and pass blocker) and credited events (e.g., player is credited with hitting the QB on the play).
In this manuscript, we use the PFF data to identify 36,362 unique players designated in the "pass rush" role across all plays.
For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e. 4-man rush formation) as the most common value.
We also leverage the PFF scouting data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
Additionally, we use the PFF player roles to identify the blocking matchup for each pass rusher in order to adjust for opponent strength, as discussed in Section \ref{sec:methods}.

>>>>

Along with the tracking information, the 2023 BDB includes scouting data provided by Pro Football Focus (PFF). 
This contains manually-collected player-level information, such as the player's role (e.g., whether they are a pass rusher and pass blocker) and credited events (e.g., player is credited with hitting the QB on the play).
In this manuscript, we use the PFF data to identify 36,362 unique players designated in the "pass rush" role across all plays.
For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e. 4-man rush formation) as the most common value.
We also leverage the PFF scouting data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
Additionally, we use the PFF player roles to identify the blocking matchup for each pass rusher in order to adjust for opponent strength, as discussed in Section \ref{sec:methods}.

### 2023-05-15 paper/03-methods.Rmd
more minor edits

## Multilevel Model for Play-Level STRAIN

@@

While average STRAIN is a simple first step for evaluating the pass rush, we can also measure pass rusher $j$'s observed average STRAIN on a single play $i$ consisting of $T_i$ total frames,

\begin{equation}
\overline{\text{STRAIN}}_{ij} = \frac{1}{T_i} \sum_{t=1}^{T_i} \widehat{\text{STRAIN}}_{ij}(t).
\end{equation}

However, the observed average STRAIN on a single play is likely due to a convolution of factors.
Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
Additionally, based on previous work by PFF, the opposing QB plays an important role on observed pressure in passing plays [@eager2019pressure].
Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers linked to a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
The coefficients $\boldsymbol{\beta}$ for additional information vector of information  $\mathbf{x_{ij}}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \cdot \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation}

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e., the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language]. Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution via parametric bootstrapping with the `merTools` package [@knowlesmerTools, @gelman2006data].

>>>>

## Multilevel Model for Play-Level STRAIN {#sec:multilevel}

@@

<!-- While average STRAIN is a simple first step for evaluating the pass rush,  -->

In addition to the average STRAIN over all frames played, we can also calculate pass rusher $j$'s observed average STRAIN on a single play $i$ consisting of $T_i$ total frames,
\begin{equation}
\overline{\text{STRAIN}}_{ij} = \frac{1}{T_i} \sum_{t=1}^{T_i} \widehat{\text{STRAIN}}_{ij}(t).
\end{equation}
While this aggregated measure is a simple first step for pass rush evaluation, the observed average STRAIN on a single play is likely due to a convolution of factors.
Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
Additionally, based on previous work by PFF, the opposing QB plays an important role on the observed pressure in passing plays [@eager2019pressure].
Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers matching up with a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
The coefficients $\boldsymbol \beta$ for covariate vector $\mathbf{x_{ij}}$ for pass rusher $j$ in play $i$ are estimated as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \cdot \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation}

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e., the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language]. Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution via parametric bootstrapping with the `merTools` package [@knowlesmerTools; @gelman2006data].

### 2023-05-15 paper/04-results.Rmd
more minor edits

In contrast, Figure \ref{fig:fig_crosby_curves_unsucess} shows the feature curves for an unsuccessful pass rush attempt by the Raiders defense from the same game.
In this play, Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass[^unsuccess].
We see that the STRAIN generated by Crosby is relatively small over the course of this play, compared to the previous play shown in Figure \ref{fig:fig_crosby_curves} which results in a sack.

@@

There are a variety of quantities we are interested in from the multilevel model fit mentioned in Section \ref{sec:methods}.
First, Table \ref{tab:fixedeff} displays the results for the fixed effects terms.
It appears that that the average STRAIN decreases by 0.042 (95% CI [-0.044, -0.040]) for every extra second in play length, and by 0.064 (95% CI [-0.069, -0.059]) every additional blocker in the offensive unit, after accounting for other covariates in the model.
In other words, NFL pass rushers tend to generate more pressure in plays with shorter length and when facing fewer number of blockers, which totally makes sense.

We observe statistically significant differences among the pass rush and pass block positions in most cases, while controlling for other variables.
The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DT and NT) appear to generate less STRAIN (relative to DE) than those that are further back, which is consistent with our results in Section \ref{sec:positional}.
For blockers, centers are those that absorb the most pressure on average, more than guards and other offensive positions (compared to the baseline level, tackles).

>>>>

In contrast, Figure \ref{fig:fig_crosby_curves_unsucess} shows the feature curves for an unsuccessful pass rush attempt by the Raiders defense from the same game.
In this play, Broncos QB Teddy Bridgewater is well-protected by the offensive line and is able to release a long pass[^unsuccess] to a receiver.
We see that the STRAIN generated by Crosby is relatively small over the course of this play, compared to the previous play shown in Figure \ref{fig:fig_crosby_curves} which results in a sack.

@@

The results of fitting the multilevel model described in Section \ref{sec:multilevel} is displayed in Table \ref{tab:fixedeff}.
First, we investigate the fixed-effects terms of this model.
It appears that that the average STRAIN decreases by 0.042 (95% CI [-0.044, -0.040]) for every extra second in play length, and by 0.064 (95% CI [-0.069, -0.059]) every additional blocker in the offensive unit, after accounting for other covariates.
In other words, NFL pass rushers tend to generate more pressure in plays with shorter length and when facing fewer number of blockers, which makes intuitive sense.

Moreover, we observe statistically significant differences among the pass rush and pass block positions in most cases, while controlling for other variables.
The coefficient estimates for the pass rush position terms reveal that players with positions closer to the line of scrimmage (DT and NT) appear to generate less STRAIN (relative to DE) than those lining up further back, which is consistent with our results in Section \ref{sec:positional}.
For blockers, centers are those that absorb the most pressure on average, more than guards and other offensive positions (compared to the baseline level, tackles).
